Accurate Binding of Sodium and Calcium to a POPC Bilayer by Effective Inclusion of Electronic Polarization

### 2018-03-16 Manuscript/SI.tex
Almost ready for second submission

\newpage
\section{Electronic continuum correction applied to CHARMM36 model}
To demonstrate that the ECC method can also be applied to other models than Lipid14,
we scaled the charges and Lennard-Jones radii of headgroup, glycerol backbone and
carbonyl atoms of the CHARMM36 model of POPC~\cite{klauda10}. We used the
same scaling factors optimized for Lipid14 in the main text, i.e., 0.8 for charges and 0.89 for Lennard-Jones
radii. In Figs. \ref{scaledCHARMM} and \ref{scaledCHARMMdensity} we show
changes of the headgroup order parameters as functions of added ions
and number density profiles along membrane normal for our ECC-CHARMM model,
together with the CHARMM36 model \cite{catte16,charmmPOPC450mMCaClfiles}
and experiments \cite{akutsu81,altenbach84}. As shown previously \cite{catte16},
the binding of Ca$^{2+}$ is significantly overestimated in CHARMM36 simulations
where no explicit repulsive Ca$^{2+}$--oxygen Lennard-Jones parameters are included.
In ECC-CHARMM simulations, the Ca$^{2+}$ binding affinity 
is reduced substantially (Fig. \ref{scaledCHARMMdensity}), and the headgroup order parameter response to the added CaCl$_2$
is in a better agreement with experiments than for the original CHARMM36 model (Fig. \ref{scaledCHARMM}).
Fig. \ref{scaledCHARMMcontacts} also shows faster equilibration times
for Ca$^{2+}$ binding in the ECC-CHARMM model than in the original CHARMM36.
Both results are in line with the findings described in the main text.

In conclusion, applying ECC to CHARMM36 simulation
improved the ion binding affinity to a lipid bilayer, despite we
directly applied the scaling factors optimized for Lipid14 without further optimizations. Also, the area per molecule in
the ECC-CHARMM simulation without ions (61 $\AA^2$) is only slightly
lower than the experimental value (64.3 $\AA^2$) \cite{kucerka11}. 
The optimal scaling factors for CHARMM36 are expected to be somewhat different
than for Lipid14 because the initial partial charges are not the same
in different models. Our results indicate that the performance of ion
binding in ECC-CHARMM36, and probably in other lipid models as well, could
be further improved by finding the optimal values for the partial
charges and Lennard-Jones radii.

It should be noted that the CHARMM36 simulation used here uses the original paramter
as in our previous study \cite{catte16,charmmPOPC450mMCaClfiles}.
Currently (February 2018), specific interaction parameters
between calcium and oxygen \cite{kim16} have been introduced in CHARMM-GUI
to reduce the binding of calcium to lipid bilayers. The impact of this changes that
dramatically reduce the calcium binding to levels similar to sodium is beyond the
scope of this work.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OP.pdf}
  \caption{\label{scaledCHARMM}
    Changes of the headgroup order parameters as a function of NaCl (left) and CaCl$_2$ (right)
    from simulations with CHARMM36 \cite{klauda10}, simulations with ECC-CHARMM parameters and experiments \cite{akutsu81,altenbach84}. 
    CHARMM36 force field parameters for simulations with and without NaCl were taken from CHARMM-gui \cite{lee15}.
    These were ran 320 ns and 280 ns, respectively, and the first 100 ns was omitted from the analysis.
    Results for CHARMM36 with CaCl$_2$ were calculated from the last 100 ns from the data available at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
    ECC-CHARMM simulations with NaCl and CaCl$_2$ were ran 260 ns and 600 ns, respectively, and the first 100 ns were omitted from the analysis.
    Sodium, calcium, and chloride ions were modeled as ECC-ions in these simulations with parameters from Refs. \citenum{martinek17, Pluharova2014}.
    All the other simulation parameters in simulations in this figure were equal to the available dataset at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
    The temperature in all simulations was set to 310 K.
    The simulation data is available at Refs. \citenum{charmm36popcT310K,charmm36popc940mMNaClT310K,ECCcharmm36popcT310K,ECCcharmm36popc940mMNaClT310K,ECCcharmm36popc450mMCaClT310K}.
  }
\end{figure}

\newpage
\begin{figure}[!h]
  \centering
  \includegraphics[width=10.0cm]{../Fig/density2.pdf}
  \caption{\label{scaledCHARMMdensity}
    Number density profiles along membrane normal from CHARMM36 \cite{klauda10} and ECC-CHARMM models with NaCl and CaCl$_2$.
    The values for the relative surface excess are also shown.
  }
\end{figure}

>>>>

\newpage
\section{Electronic continuum correction applied to CHARMM36 model}
To demonstrate that the ECC method can also be applied to other models than Lipid14,
we scaled the charges and Lennard-Jones radii of headgroup, glycerol backbone and
carbonyl atoms of the CHARMM36 model of POPC~\cite{klauda10}. We used the
same scaling factors optimized for Lipid14 in the main text, i.e., 0.8 for charges and 0.89 for Lennard-Jones
radii. In Figs. \ref{scaledCHARMM} and \ref{scaledCHARMMdensity} we show
changes of the headgroup order parameters as functions of added ions
and number density profiles along membrane normal for our ECC-CHARMM36 model,
together with the CHARMM36 model \cite{catte16,charmmPOPC450mMCaClfiles}
and experiments \cite{akutsu81,altenbach84}. As shown previously \cite{catte16},
the binding of Ca$^{2+}$ is significantly overestimated in CHARMM36 simulations
where no explicit repulsive Ca$^{2+}$--oxygen Lennard-Jones parameters are included.
In ECC-CHARMM36 simulations, the Ca$^{2+}$ binding affinity 
is reduced substantially (Fig. \ref{scaledCHARMMdensity}), and the headgroup order parameter response to the added CaCl$_2$
is in a better agreement with experiments than for the original CHARMM36 model (Fig. \ref{scaledCHARMM}).
Fig. \ref{scaledCHARMMcontacts} also shows faster equilibration times
for Ca$^{2+}$ binding in the ECC-CHARMM36 model than in the original CHARMM36.
Both results are in line with the findings described in the main text.

In conclusion, applying ECC to CHARMM36 simulation
improved the ion binding affinity to a lipid bilayer, despite we
directly applied the scaling factors optimized for Lipid14 without further optimizations. Also, the area per molecule in
the ECC-CHARMM36 simulation without ions (61 $\AA^2$) is only slightly
lower than the experimental value (64.3 $\AA^2$) \cite{kucerka11}. 
The optimal scaling factors for CHARMM36 are expected to be somewhat different
than for Lipid14 because the initial partial charges are not the same
in the two models. Our results indicate that the performance of ion
binding in ECC-CHARMM36, and probably in other lipid models as well, could
be further improved by finding the optimal values for the partial
charges and Lennard-Jones radii.

It should be noted that the CHARMM36 simulation used here uses the original paramter
as in our previous study \cite{catte16,charmmPOPC450mMCaClfiles}.
Currently (February 2018), specific interaction parameters
between calcium and oxygen \cite{kim16} have been introduced in CHARMM-GUI \cite{lee15}
to reduce the binding of calcium to lipid bilayers. The further discussion of these
changes that dramatically reduce the calcium binding to levels similar to sodium is beyond the
scope of this work.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OP.pdf}
  \caption{\label{scaledCHARMM}
    Changes of the headgroup order parameters as a function of NaCl (left) and CaCl$_2$ (right)
    from simulations with CHARMM36 \cite{klauda10}, simulations with ECC-CHARMM36 parameters and experiments \cite{akutsu81,altenbach84}. 
    CHARMM36 force field parameters for simulations with and without NaCl were taken from CHARMM-GUI \cite{lee15}.
    These were ran 320 ns and 280 ns, respectively, and the first 100 ns was omitted from the analysis.
    Results for CHARMM36 with CaCl$_2$ were calculated from the last 100 ns from the data available at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
    ECC-CHARMM36 simulations with NaCl and CaCl$_2$ were ran 260 ns and 600 ns, respectively, and the first 100 ns were omitted from the analysis.
    Sodium, calcium, and chloride ions were modeled as ECC-ions in these simulations with parameters from Refs. \citenum{martinek17, Pluharova2014}.
    All the other simulation parameters in simulations in this figure were equal to the available dataset at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
    The temperature in all simulations was set to 310 K.
    The simulation data is available at Refs. \citenum{charmm36popcT310K,charmm36popc940mMNaClT310K,ECCcharmm36popcT310K,ECCcharmm36popc940mMNaClT310K,ECCcharmm36popc450mMCaClT310K}.
  }
\end{figure}

\newpage
\begin{figure}[!h]
  \centering
  \includegraphics[width=10.0cm]{../Fig/density2.pdf}
  \caption{\label{scaledCHARMMdensity}
    Number density profiles along membrane normal from CHARMM36 \cite{klauda10} and ECC-CHARMM36 models with NaCl and CaCl$_2$.
    The values for the relative surface excess are also shown.
  }
\end{figure}

### 2018-03-16 Manuscript/manuscript.tex
Almost ready for second submission

Here we develop a ECC-POPC lipid model that accurately describes binding 
of sodium and calcium ions to the POPC  lipid bilayer. 
The Lipid14~\cite{dickson14} force field 
(available in a Gromacs format from Ref. \citenum{lipid14files}) was used as a starting 
point since 
it provides the most accurate response of the head groups to ions among the available 
lipid models (see Figs.~2~and~5 in Ref.~\citenum{catte16}). Additionally, the Lipid14 model 
provides relatively realistic head group, glycerol backbone, and acyl chain structures~\cite{dickson14,botan15}.
We applied the ECC correction 
to the Lipid14 model of POPC by scaling 
partial charges of the head group, glycerol 
backbone, and carbonyl regions. 
These are the polar parts of phospholipids which can 
contribute to the cation binding.
A similar approach is also applied to the CHARMM36 model of POPC~\cite{klauda10} in SI.

@@

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
The topology of the dihexadecyldimethylammonium cation was created with the automated topology builder \cite{malde11}.
The AmberTools program~\cite{amber} was used to generate the Amber-type force field parameters. These parameters were
then converted to the Gromacs format with the acpype tool \cite{acpype}. The partial charges were manually modified in
the non-scaled version to approximately match similar segments in Lipid14 \cite{dickson14}.
For simulations of cations mixed with ECC-POPC, the ECC version of cation force field was made
by scaling the head group parameters with the same scaling factors that were
used for ECC-POPC, i.e., $f_q=0.8$ for charges and 0.89 for $\sigma$.

@@

The results from simulations combining the ECC-POPC with the ECC-ion models \cite{martinek17, kohagen16, Pluharova2014} exhibit a significantly improved behavior of the POPC head group order parameters as a function of NaCl or CaCl$_2$ concentrations, see Fig.~\ref{fig:delta_ordPar_NaCl} and Fig.~\ref{fig:delta_ordPar_CaCl}. Considering that we are also able to reproduce the experimental response in systems with known charge density (see above section \ref{section:boundCHARGE}), we conclude that our ECC model correctly reproduces the binding affinities of Na$^{+}$ and Ca$^{2+}$ ions to the POPC lipid bilayer.
Also the binding affinity of K$^+$ ions to POPC bilayer is in line with the experimental data \cite{eisenberg79,binder02,klasczyk10}, as shown in SI.
Furthermore, while the response of the glycerol backbone $g_3$ order parameter to CaCl$_2$ was significantly overestimated in the original Lipid14 model, the ECC-POPC model provides an improved agreement with experiment, as seen in Fig.~\ref{fig:delta_ordPar_CaCl}.
Also the changes of the P-N vector angle are too pronounced for the Lipid14 model,
for which the largest tilting toward water phase induced by a $780\,\mathrm{mM}$
CaCl$_2$ concentration is approximately 17$^{\circ}$. The corresponding value
for the ECC-POPC simulation is only 6$^{\circ}$ ($820\,\mathrm{mM}$ CaCl$_2$).

@@

In this study, we employed the electrometer concept to demonstrate that the binding of \ce{Na^+} and \ce{Ca^2+} ions to a POPC lipid bilayer can be accurately described within a force field MD simulation, provided that electronic polarization is implicitly included via the electronic continuum correction~\cite{leontyev11}. This ECC-POPC model is built upon the Lipid14 POPC model~\cite{dickson14} by scaling the partial charges by a factor of 0.8 and reducing the Lennard-Jones radii by a factor of 0.89 for the headgroup, glycerol backbone, and carbonyl atoms. While the structural details of a POPC lipid bilayer in pure water simulated with the newly developed ECC-POPC model agree with experiments with an accuracy comparable to the other state of the art lipid models, the new model in addition reproduces the experimental lipid head group order parameter responses to a cationic surfactant, NaCl, and CaCl$_2$ at varying concentrations. It thus represents a significant improvement over currently available lipid models, which tend to overestimate cation binding affinities~\cite{catte16}.
A similar improvement using the ECC procedure is demonstrated for the POPC model of CHARMM36 \cite{klauda10} in SI.

>>>>

Here we develop a ECC-POPC lipid model that accurately describes binding 
of sodium and calcium ions to the POPC  lipid bilayer. 
The Lipid14~\cite{dickson14} force field 
(available in a Gromacs format from Ref. \citenum{lipid14files}) was used as a starting 
point since 
it provides the most accurate response of the head groups to ions among the available 
lipid models (see Figs.~2~and~5 in Ref.~\citenum{catte16}). Additionally, the Lipid14 model 
provides relatively realistic head group, glycerol backbone, and acyl chain structures~\cite{dickson14,botan15}.
We applied the ECC correction 
to the Lipid14 model of POPC by scaling 
partial charges of the head group, glycerol 
backbone, and carbonyl regions. 
These are the polar parts of phospholipids which can 
contribute to the cation binding.
A similar approach is also applied to the CHARMM36 model of POPC~\cite{klauda10} in the the SI.

@@

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
The topology of the dihexadecyldimethylammonium cation was created with the automated topology builder \cite{malde11}.
The AmberTools program~\cite{amber} was used to generate the Amber-type force field parameters. These parameters were
then converted to the Gromacs format with the acpype tool \cite{acpype}. The partial charges were manually modified in
the non-scaled version to approximately match similar segments in Lipid14 \cite{dickson14}.
For simulations of cationic surfactants mixed with ECC-POPC,
all charges of cationic surfactants were scaled by a factor $f_q$=0.75
consistently with the counterions. The Lennard-Jones radii were scaled with
the same factor as for ECC-POPC, i.e. $f_{\sigma}$=0.89.

%the ECC version of the cationic surgactnat force field was made
%by scaling the partial charges and Lennard-Jones parameters of the head group parameters with the same scaling factors as in
%ECC-POPC, i.e., $f_q=0.8$ for charges and 0.89 for $\sigma$.

%For simulations of THE cations mixed with ECC-POPC,

@@

The results from simulations combining the ECC-POPC with the ECC-ion models \cite{martinek17, kohagen16, Pluharova2014} exhibit a significantly improved behavior of the POPC head group order parameters as a function of NaCl or CaCl$_2$ concentrations, see Fig.~\ref{fig:delta_ordPar_NaCl} and Fig.~\ref{fig:delta_ordPar_CaCl}. Considering that we are also able to reproduce the experimental response in systems with known charge density (see above section \ref{section:boundCHARGE}), we conclude that our ECC model correctly reproduces the binding affinities of Na$^{+}$ and Ca$^{2+}$ ions to the POPC lipid bilayer.
Also the binding affinity of K$^+$ ions to POPC bilayer is in line with the experimental data \cite{eisenberg79,binder02,klasczyk10} (shown in SI).
Furthermore, while the response of the glycerol backbone $g_3$ order parameter to CaCl$_2$ was significantly overestimated in the original Lipid14 model, the ECC-POPC model provides an improved agreement with experiment, as seen in Fig.~\ref{fig:delta_ordPar_CaCl}.
Also the changes of the P-N vector angle are too pronounced for the Lipid14 model,
for which the largest tilting toward water phase induced by a $780\,\mathrm{mM}$
CaCl$_2$ concentration is approximately 17$^{\circ}$. The corresponding value
for the ECC-POPC simulation is only 6$^{\circ}$ ($820\,\mathrm{mM}$ CaCl$_2$).

@@

In this study, we employed the electrometer concept to demonstrate that the binding of \ce{Na^+} and \ce{Ca^2+} ions to a POPC lipid bilayer can be accurately described within a force field MD simulation, provided that electronic polarization is implicitly included via the electronic continuum correction~\cite{leontyev11}. This ECC-POPC model is built upon the Lipid14 POPC model~\cite{dickson14} by scaling the partial charges by a factor of 0.8 and reducing the Lennard-Jones radii by a factor of 0.89 for the headgroup, glycerol backbone, and carbonyl atoms. While the structural details of a POPC lipid bilayer in pure water simulated with the newly developed ECC-POPC model agree with experiments with an accuracy comparable to the other state of the art lipid models, the new model in addition reproduces the experimental lipid head group order parameter responses to a cationic surfactant, NaCl, and CaCl$_2$ at varying concentrations. It thus represents a significant improvement over currently available lipid models, which tend to overestimate cation binding affinities~\cite{catte16}.
A similar improvement using the ECC procedure is demonstrated for the POPC model of CHARMM36 \cite{klauda10} in the SI.
